---
title: "BTS Flight Time Regression Results -- 1 year"
output: 
  html_document:
    self_contained: true
    toc: true
    toc_depth: 1
    toc_float:
      collapsed: true
      smooth_scroll: false
---

<!-- saved from url=(0014)about:internet -->
<!-- This needs to be manually added to the second line of the html -->

# Overview 

Summary of results from models of all reportable flights in 2018. 

The modeling framework is as follows:
$$
AIRTIME_{CARRIER_{O-D}} = \beta1_{CARRIER_{O-D}} MONTH_{CARRIER_{O-D}} + \beta2_{CARRIER_{O-D}}DAY_{CARRIER_{O-D}} + \beta3_{CARRIER_{O-D}}DEP_{CARRIER_{O-D}}
$$
For each carrier, each origin-destination flight segment was modeled as a linear combination of month of year, day of week, and departure time block (hour of day). These are the variables that an air carrier would know when scheduling a flight, and thus what would be available to provide a prediction of estimated flight time for a consumer at the time of booking. Two- and three-way interactions were included if they improved model fit by AIC.

Additional variables to consider could be predictable climatological features (El Nino) or expected airport congestion. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning=F, message=F)

library(tidyverse)
library(knitr)
library(DT)
library(kableExtra)
library(plotly) # note: masks filter() from dplyr

codeloc = ifelse(grepl('Flynn', normalizePath('~/')),
                   "~/git/Wheels_Up",
                   "~/GitHub/Wheels_Up")
sharedloc = "//vntscex.local/DFS/Projects/PROJ-OR02A2/SDI/BTS_Flight_performance/Data"
resultsloc = "//vntscex.local/DFS/Projects/PROJ-OR02A2/SDI/BTS_Flight_performance/Results"

```

TODO:

- Summarize n flight segments, r2 distribution by carrier

- Summarize n flights, r2 distribution by origin and destination

- R2 by N, R2 by mean flight time, and r2 by sd flight time plots

- Table of most and least predictable flights

# Summary tables 

The following table summarizes the predictability of flights by carrier. For each carrier, the number of unique flight segments (origin-destination pairs) is provided. Not all flight segments could be reliably modeled; for instance, American flew only a single flight between DCA and Atlanta in 2018 (on 2018-12-23, at 9pm). With only a single value for this combination of carrier and origin-destination pair, the current modeling framework does not produce an estimate for this flight segment.

A hierarchical mixed effect model would produce an estimate for such a flight segment, by pooling information from other carrier flying the same route. However, we found these models to have difficulty converging using a Monte Carlo Markov Chain approach, so have elected to use the ordinary least squares regression approach, meaning each carrier x origin-destination pair is modeled separately.


```{r carr_fl_prep}
ols_res <- dir(file.path(resultsloc, 'OLS_2018'))

d <- vector()
for(i in 1:length(ols_res)){
  dx <- read.csv(file.path(resultsloc, 'OLS_2018', ols_res[i]))
  carrier = substr(ols_res[i], 1, 2)
  dx <- data.frame(carrier, dx)
  dx <- dx %>%
    filter(O_D != '') %>%
    mutate(Origin = substr(O_D, 1, 3),
           Destination = substr(O_D, 5, 7))
  
  d <- rbind(d, dx)
}
```



```{r carr_summary}
d_carrier <- d %>%
  group_by(carrier) %>%
  summarize(N_flight_segments = n(),
            N_estimated = sum(!is.na(r2)),
            R2_median = round(median(r2, na.rm=T), 3),
            R2_max = round(max(r2[r2 != 1], na.rm = T), 3),
            Pct_High_R2 = round(100*sum(!is.na(r2) & r2 > 0.5) / length(r2), 2))

kable(d_carrier,
      format.args = list(big.mark = ','),
      caption = "Summary of flight predictability by carrier in 2018." ) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))

```



```{r airport_summary}
d_airport <- d %>%
  group_by(Origin) %>%
  summarize(N_flight_segments = n(),
            N_estimated = sum(!is.na(r2)),
            R2_median = round(median(r2, na.rm=T), 3),
            R2_max = round(max(r2[r2 != 1], na.rm = T), 3),
            Pct_High_R2 = round(100*sum(!is.na(r2) & r2 > 0.5) / length(r2), 2))

datatable(d_airport,
          rownames = F,
          filter = 'top',
          options = list(order = list(list(3, 'desc')),
                        searchCols = list(NULL, list(search = '100 ... 1000'), NULL, NULL, NULL, NULL, NULL),
                         pageLength = 10),
      caption = "Summary of flight predictability by origin airport in 2018. Airports with at least 100 departing flight segments shown by default." ) 

# TODO: load full airport names for hover over
```