---
title: "BTS Flight Time Regression Results"
output: 
  html_document:
    self_contained: true
    toc: true
    toc_depth: 1
    toc_float:
      collapsed: true
      smooth_scroll: false
---

<!-- saved from url=(0014)about:internet -->
<!-- This needs to be manually added to the second line of the html -->

# Overview 

Summary of results from models of all reportable flights from 2015 - 2018. 

Two modeling approaches were taken. First, individually modeling each flight segment (origin-destination, O-D) by each carrier (C):

$$
AIRTIME_{C_{O-D}} = \beta1_{C_{O-D}} Year_{C_{O-D}} + \beta2_{C_{O-D}} MONTH_{C_{O-D}} + \beta3_{C_{O-D}}DAY_{C_{O-D}} + \beta4_{C_{O-D}}DEP_{C_{O-D}}
$$

For each carrier, each origin-destination flight segment was modeled as a linear combination of month of year, day of week, and departure time block (hour of day). These are the variables that an air carrier would know when scheduling a flight, and thus what would be available to provide a prediction of estimated flight time for a consumer at the time of booking. Two- and three-way interactions were included if they improved model fit by AIC. The best fit model varied by flight segment and carrier, with inclusion of interactions determined by AIC.

Second, individually modeling each flight segment (origin-destination, O-D) across carrier:

$$
AIRTIME_{O-D} = \beta1_{O-D} CARRIER_{O-D} + \beta2_{O-D} YEAR_{O-D} + \beta3_{O-D} MONTH_{O-D} + \beta4_{O-D}DAY_{O-D} + \beta5_{O-D}DEP_{O-D}
$$

This approach differs from the first by including all flights on a particular flight segment regardless of carrier. The carrier-specific effects are then an offset from the overall estiamte, with two- and three-way interactions again included to the extent that they improved model fit. 

Additional variables to consider could be predictable climatological features (El Nino) or expected airport congestion. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning=F, message=F)

library(tidyverse)
library(knitr)
library(DT)
library(kableExtra)
library(plotly) # note: masks filter() from dplyr

codeloc = ifelse(grepl('Flynn', normalizePath('~/')),
                   "~/git/Wheels_Up",
                   "~/GitHub/Wheels_Up")
sharedloc = "//vntscex.local/DFS/Projects/PROJ-OR02A2/SDI/BTS_Flight_performance/Data"
resultsloc = "//vntscex.local/DFS/Projects/PROJ-OR02A2/SDI/BTS_Flight_performance/Results"

```

TODO:

- Summarize n flight segments, r2 distribution by carrier

- Summarize n flights, r2 distribution by origin and destination

- R2 by N, R2 by mean flight time, and r2 by sd flight time plots

- Table of most and least predictable flights

# Summary tables 

The following table summarizes the predictability of flights by carrier. For each carrier, the number of unique flight segments (origin-destination pairs) is provided. Not all flight segments could be reliably modeled; for instance, American flew only a single flight between DCA and Atlanta in 2018 (on 2018-12-23, at 9pm). With only a single value for this combination of carrier and origin-destination pair, the current modeling framework does not produce an estimate for this flight segment.

A hierarchical mixed effect model would produce an estimate for such a flight segment, by pooling information from other carrier flying the same route. However, we found these models to have difficulty converging using a Monte Carlo Markov Chain approach, so have elected to use the ordinary least squares regression approach, meaning each carrier x origin-destination pair is modeled separately.


```{r carr_fl_prep}
ols_res <- dir(file.path(resultsloc, 'OLS_2018'))

d <- vector()
for(i in 1:length(ols_res)){
  dx <- read.csv(file.path(resultsloc, 'OLS_2018', ols_res[i]))
  carrier = substr(ols_res[i], 1, 2)
  dx <- data.frame(carrier, dx)
  dx <- dx %>%
    filter(O_D != '') %>%
    mutate(Origin = substr(O_D, 1, 3),
           Destination = substr(O_D, 5, 7))
  
  d <- rbind(d, dx)
}

d_crossyear <- read_csv(file.path(resultsloc, '1O_D_Crossyear_OLS_Summary.csv'))
d_crossyear <- d_crossyear %>%
  filter(O_D != '') %>%
  mutate(Origin = substr(O_D, 1, 3),
         Destination = substr(O_D, 5, 7))

```



```{r carr_summary}
d_carrier <- d %>%
  group_by(carrier) %>%
  summarize(N_flight_segments = n(),
            N_estimated = sum(!is.na(r2)),
            R2_median = round(median(r2, na.rm=T), 3),
            R2_max = round(max(r2[r2 != 1], na.rm = T), 3),
            Pct_High_R2 = round(100*sum(!is.na(r2) & r2 > 0.5) / length(r2), 2))

kable(d_carrier,
      format.args = list(big.mark = ','),
      caption = "Summary of flight predictability by carrier in 2018." ) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))

```

## Airport summary -- Model within carrier

```{r airport_summary}
d_airport <- d %>%
  group_by(Origin) %>%
  summarize(N_flight_segments = n(),
            N_estimated = sum(!is.na(r2)),
            R2_median = round(median(r2, na.rm=T), 3),
            R2_max = round(max(r2[r2 != 1], na.rm = T), 3),
            Pct_High_R2 = round(100*sum(!is.na(r2) & r2 > 0.5) / length(r2), 2))

datatable(d_airport,
          rownames = F,
          filter = 'top',
          options = list(order = list(list(3, 'desc')),
                        searchCols = list(NULL, list(search = '100 ... 1000'), NULL, NULL, NULL, NULL, NULL),
                         pageLength = 10),
      caption = "Summary of flight predictability by origin airport in 2018. Airports with at least 100 departing flight segments shown by default." ) 

# TODO: load full airport names for hover over
```

## Airport summary -- Model across carrier

```{r airport_summary_across}
d_airport <- d_crossyear %>%
  group_by(Origin) %>%
  summarize(N_flight_segments = n(),
            N_estimated = sum(!is.na(r2)),
            R2_median = round(median(r2, na.rm=T), 3),
            R2_max = round(max(r2[r2 != 1], na.rm = T), 3),
            Pct_High_R2 = round(100*sum(!is.na(r2) & r2 > 0.5) / length(r2), 2))

datatable(d_airport,
          rownames = F,
          filter = 'top',
          options = list(order = list(list(3, 'desc')),
                        searchCols = list(NULL, list(search = '100 ... 1000'), NULL, NULL, NULL, NULL, NULL),
                         pageLength = 10),
      caption = "Summary of flight predictability by origin airport, across years and modeled across carriers. Airports with at least 100 departing flight segments shown by default." ) 
```