---
title: "FAA Flight Delay EDA"
output: 
  html_document:
    self_contained: true
    toc: true
    toc_depth: 1
    toc_float:
      collapsed: true
      smooth_scroll: false
---

<!-- saved from url=(0014)about:internet -->
<!-- This needs to be manually added to the second line of the html -->

# Overview 

This document provides an overview of the FAA flight delay data. This document currently reviews the 2015, 2016, and 2017 Airline Service Quality Performance ([ASQP](https://aspmhelp.faa.gov/index.php/Airline_Service_Quality_Performance_(ASQP))) data; 2018 data is in process.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning=F, message=F)

library(tidyverse)
library(knitr)
library(DT)
library(kableExtra)
library(plotly) # note: masks filter() from dplyr

codeloc = ifelse(grep('Flynn', normalizePath('~/')),
                "~/git/Wheels_Up",
                "Erika_put_your_path_here/Wheels_Up")
sharedloc = "//vntscex.local/DFS/Projects/PROJ-OR02A2/SDI/BTS_Flight_performance/Data"

avail_data = dir(sharedloc)
avail_data = avail_data[grep('^ASQP.\\d{4}.RData$', avail_data)] # Match with regular expresion to only prepared .RData files

if(length(avail_data) == 0) {
    source(file.path(codeloc, 'datacleaning', 'Data_prep.R'))
} else {
    for(x in avail_data){
      load(file.path(sharedloc, x))
    }
  load(file.path(sharedloc, 'Cross-year_Summary.RData'))
  }
```

The following table summarizes the counts of flights by carrier across the three currently available years.

```{r carr_yr_table}
carr_yr_table = rbind(carr_yr, c('Total', colSums(carr_yr[2:ncol(carr_yr)], na.rm = T)))
class(carr_yr_table) = 'data.frame'
colnames(carr_yr_table) = c('Carrier', '2015', '2016', '2017')
for(i in 2:ncol(carr_yr_table)){ carr_yr_table[,i] = as.numeric(carr_yr_table[,i]) }

kable(carr_yr_table,
      format.args = list(big.mark = ','),
      caption = "Counts of flights by year and carrier in ASQP data." ) %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  row_spec(nrow(carr_yr_table), bold = T,hline_after = T)

```

## Variables {.tabset .tabset-pills}

The following tables summarize the variables being retained from the complete ASQP data. These tables allow for a quality assurance check on completeness and correctness of the data for each year.

### 2015
The data frame has `r format(nrow(d_15), big.mark=",")` rows and `r ncol(d_15)` columns.

```{r overview15}
overview_df = data.frame(Variables = names(d_15),
                         Class = sapply(d_15, class),
                         N_unique = sapply(d_15, function(x) length(unique(x))),
                         Min_numeric = sapply(d_15, function(x) ifelse(is.numeric(x), min(x, na.rm=T), "")),
                         Max_numeric = sapply(d_15, function(x) ifelse(is.numeric(x), max(x, na.rm=T), "")),
                         Top_factor = sapply(d_15, function(x) ifelse(is.factor(x), names(sort(table(x), dec=T))[1], ""))
                         )

kable(overview_df, 
       row.names = F,
       caption = "Overview of 2015 ASQP data. For factor variables, most frequent value is shown." ) %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```


### 2016
The data frame has `r format(nrow(d_16), big.mark=",")` rows and `r ncol(d_16)` columns.

```{r overview16}
overview_df = data.frame(Variables = names(d_16),
                         Class = sapply(d_16, class),
                         N_unique = sapply(d_16, function(x) length(unique(x))),
                         Min_numeric = sapply(d_16, function(x) ifelse(is.numeric(x), min(x, na.rm=T), "")),
                         Max_numeric = sapply(d_16, function(x) ifelse(is.numeric(x), max(x, na.rm=T), "")),
                         Top_factor = sapply(d_16, function(x) ifelse(is.factor(x), names(sort(table(x), dec=T))[1], ""))
                         )

kable(overview_df, 
       row.names = F,
       caption = "Overview of 2016 ASQP data. For factor variables, most frequent value is shown." ) %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```


### 2017
The data frame has `r format(nrow(d_17), big.mark=",")` rows and `r ncol(d_17)` columns.

```{r overview17}
overview_df = data.frame(Variables = names(d_17),
                         Class = sapply(d_17, class),
                         N_unique = sapply(d_17, function(x) length(unique(x))),
                         Min_numeric = sapply(d_17, function(x) ifelse(is.numeric(x), min(x, na.rm=T), "")),
                         Max_numeric = sapply(d_17, function(x) ifelse(is.numeric(x), max(x, na.rm=T), "")),
                         Top_factor = sapply(d_17, function(x) ifelse(is.factor(x), names(sort(table(x), dec=T))[1], ""))
                         )

kable(overview_df, 
       row.names = F,
       caption = "Overview of 2017 ASQP data. For factor variables, most frequent value is shown.") %>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```


# Visualization 

For each year, the following visuals provide a check in the delay data provided in ASQP. For this project, we will calculate delay based on wheels-off to wheels-on compared to the scheduled departure and arrival.

## Plotting delays by day and airline {.tabset .tabset-pills}

### 2015

```{r explore15}
delay_summary <- d_15 %>%
  group_by(CARRIER, MONTH, FLIGHT_DATE) %>%
  summarize(n_flights = n(),
            mean_dep_delay = mean(as.numeric(DEP_DELAY_MINS), na.rm=T),
            mean_arr_delay = mean(as.numeric(ARR_DELAY_MINS), na.rm=T),
            sd_dep_delay = sd(as.numeric(DEP_DELAY_MINS), na.rm=T),
            sd_arr_delay = sd(as.numeric(ARR_DELAY_MINS), na.rm=T))

delay_summary_origin <- d_15 %>%
  group_by(ORIGIN, MONTH, FLIGHT_DATE) %>%
  summarize(n_flights = n(),
            mean_dep_delay = mean(as.numeric(DEP_DELAY_MINS), na.rm=T),
            mean_arr_delay = mean(as.numeric(ARR_DELAY_MINS), na.rm=T),
            sd_dep_delay = sd(as.numeric(DEP_DELAY_MINS), na.rm=T),
            sd_arr_delay = sd(as.numeric(ARR_DELAY_MINS), na.rm=T))

busy_airports <- d_15 %>%
  group_by(ORIGIN) %>%
  summarize(n_flights = n()) %>%
  filter(n_flights >= sort(n_flights, decreasing = T)[6])

```

```{r plot15, fig.height=5, fig.width=7}
p1 <- ggplot(delay_summary) +
 geom_point(aes(x = FLIGHT_DATE, y = mean_arr_delay, color = CARRIER,
                text = paste("N:", n_flights))) +
  ggtitle("Mean arrival delay in minutes by carrier for 2015")

ggplotly(p1)

p2 <- ggplot(delay_summary_origin %>% filter(n_flights > 250 & !is.na(mean_arr_delay) & ORIGIN %in% busy_airports$ORIGIN)) +
  geom_histogram(aes(mean_arr_delay, fill = ORIGIN)) +
  facet_wrap(~ORIGIN, ncol = 3) +
  ggtitle('Distribution of mean daily arrival delay', subtitle = 'For days with >250 flights in the top 6 busiest airports')
ggplotly(p2)
```


### 2016

```{r explore16}
delay_summary <- d_16 %>%
  group_by(CARRIER, MONTH, FLIGHT_DATE) %>%
  summarize(n_flights = n(),
            mean_dep_delay = mean(as.numeric(DEP_DELAY_MINS), na.rm=T),
            mean_arr_delay = mean(as.numeric(ARR_DELAY_MINS), na.rm=T),
            sd_dep_delay = sd(as.numeric(DEP_DELAY_MINS), na.rm=T),
            sd_arr_delay = sd(as.numeric(ARR_DELAY_MINS), na.rm=T))

delay_summary_origin <- d_16 %>%
  group_by(ORIGIN, MONTH, FLIGHT_DATE) %>%
  summarize(n_flights = n(),
            mean_dep_delay = mean(as.numeric(DEP_DELAY_MINS), na.rm=T),
            mean_arr_delay = mean(as.numeric(ARR_DELAY_MINS), na.rm=T),
            sd_dep_delay = sd(as.numeric(DEP_DELAY_MINS), na.rm=T),
            sd_arr_delay = sd(as.numeric(ARR_DELAY_MINS), na.rm=T))

busy_airports <- d_16 %>%
  group_by(ORIGIN) %>%
  summarize(n_flights = n()) %>%
  filter(n_flights >= sort(n_flights, decreasing = T)[6])

```

```{r plot16, fig.height=5, fig.width=7}
p1 <- ggplot(delay_summary) +
 geom_point(aes(x = FLIGHT_DATE, y = mean_arr_delay, color = CARRIER,
                text = paste("N:", n_flights))) +
  ggtitle("Mean arrival delay in minutes by carrier for 2016")

ggplotly(p1)

p2 <- ggplot(delay_summary_origin %>% filter(n_flights > 250 & !is.na(mean_arr_delay) & ORIGIN %in% busy_airports$ORIGIN)) +
  geom_histogram(aes(mean_arr_delay, fill = ORIGIN)) +
  facet_wrap(~ORIGIN, ncol = 3) +
  ggtitle('Distribution of mean daily arrival delay', subtitle = 'For days with >250 flights in the top 6 busiest airports')
ggplotly(p2)
```

### 2017

```{r explore17}
delay_summary <- d_17 %>%
  group_by(CARRIER, MONTH, FLIGHT_DATE) %>%
  summarize(n_flights = n(),
            mean_dep_delay = mean(as.numeric(DEP_DELAY_MINS), na.rm=T),
            mean_arr_delay = mean(as.numeric(ARR_DELAY_MINS), na.rm=T),
            sd_dep_delay = sd(as.numeric(DEP_DELAY_MINS), na.rm=T),
            sd_arr_delay = sd(as.numeric(ARR_DELAY_MINS), na.rm=T))

delay_summary_origin <- d_17 %>%
  group_by(ORIGIN, MONTH, FLIGHT_DATE) %>%
  summarize(n_flights = n(),
            mean_dep_delay = mean(as.numeric(DEP_DELAY_MINS), na.rm=T),
            mean_arr_delay = mean(as.numeric(ARR_DELAY_MINS), na.rm=T),
            sd_dep_delay = sd(as.numeric(DEP_DELAY_MINS), na.rm=T),
            sd_arr_delay = sd(as.numeric(ARR_DELAY_MINS), na.rm=T))

busy_airports <- d_17 %>%
  group_by(ORIGIN) %>%
  summarize(n_flights = n()) %>%
  filter(n_flights >= sort(n_flights, decreasing = T)[6])

```

```{r plot17, fig.height=5, fig.width=7}
p1 <- ggplot(delay_summary) +
 geom_point(aes(x = FLIGHT_DATE, y = mean_arr_delay, color = CARRIER,
                text = paste("N:", n_flights))) +
  ggtitle("Mean arrival delay in minutes by carrier for 2017")

ggplotly(p1)

p2 <- ggplot(delay_summary_origin %>% filter(n_flights > 250 & !is.na(mean_arr_delay) & ORIGIN %in% busy_airports$ORIGIN)) +
  geom_histogram(aes(mean_arr_delay, fill = ORIGIN)) +
  facet_wrap(~ORIGIN, ncol = 3) +
  ggtitle('Distribution of mean daily arrival delay', subtitle = 'For days with >250 flights in the top 6 busiest airports')
ggplotly(p2)
```

# Wheels off/on time compared

Confirming: 

```
  WHEELS_OFF = DEP_TIME + TAXI_OUT
  WHEELS_ON  = ARR_TIME - TAXI_IN
```

```{r, wheels_time}
kable(head(d_15 %>% select(DEP_TIME_HR, DEP_TIME_MIN, TAXI_OUT, WHEELS_OFF, WHEELS_ON, TAXI_IN, ARR_TIME_HR, ARR_TIME_MIN)),
      caption = 'Example DEP, ARR, TAXI, and WHEELS time') %>% 
    kable_styling(bootstrap_options = c("striped", "hover"))
```

Scheduled time:

- CRS = Computerized Reservation System

```{r crs_time}
kable(head(d_15 %>% select(CRS_DEP_TIME_HR, CRS_DEP_TIME_MIN, WHEELS_OFF, WHEELS_ON, CRS_ARR_TIME_HR, CRS_ARR_TIME_MIN)),
      caption = 'Example WHEELS and CRS time') %>%
      kable_styling(bootstrap_options = c("striped", "hover"))

```

# Notes

Questions:

- NULL values for `WHEELS_OFF` or `WHEELS_ON` indicate no value in input data. What reasons might lead to missing values?
- CRS time is the Computerized Reservation System. Should we take CRS time as the scheduled time that a consumer would see 'at booking'?

Data processing notes:

- WAC (World Area Code) is synonymous with state, we have omitted it to save space.
- Have kept ORIGIN_CITY_NAME because assume weather events at one city will be shared across airports

Additional data:

- Table of lat / long for each airport. Obtained from BTS [here](https://www.transtats.bts.gov/tables.asp?Table_ID=315&SYS_Table_Name=T_WAC_COUNTRY_STATE), will update data after integrating.
- Time zone for departure and arrival airports. We will want this so that we can do our own calculations of time differences; we can generate this given lat longs of each airport if needed. Now that we have lat long, can overlay with timezone shapefile.


## Merger of AA and US

Counts of flights by month for 2015, by carrier, show the effect of the merger between AA and US Airways in 2015 ([AA/US merger](https://en.wikipedia.org/wiki/American_Airlines_Group); [Airline codes here](https://aspmhelp.faa.gov/index.php/ASQP:_Carrier_Codes_and_Names)). Other years do not show any similar breaks in count of flights by month and carrier.

```{r summary2}
kable(table(d_15$CARRIER, d_15$MONTH),
      caption = 'Count of flights by month of 2015, by carrier',
      format.args = list(big.mark = ',')) %>%
    kable_styling(bootstrap_options = c("striped", "hover"))


# kable(table(d_17$CARRIER, d_17$MONTH),
#       caption = 'Count of flights by month of 2017, by carrier',
#       format.args = list(big.mark = ',')) %>%
#     kable_styling(bootstrap_options = c("striped", "hover"))

```